19Q1

### 2019-05-01 19Q1.Rmd
updated Rmd

    # Get all G2Pgene-cell line combinations
    grid <- expand.grid("Cell_line" = unique(filt$Cell_line), "Gene" = g2p_genes)
    data <-  merge(filt, grid, by = "Cell_line", all.y = TRUE)
    print("Merged grid.")
    # Merge metadata
    data <-  merge(data, select(ccl_meta, Cell_line, Histology, CCLE_Name, Broad_ID, Primary.Disease), by = c("Cell_line", "Histology", "CCLE_Name", "Broad_ID"), all.x = TRUE)
    ## Add columns for G2P drug-gene and drug-gene-lineage sorting
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$G2P_Drug_Gene <- ifelse(data$Drug_Gene %in% g2p_druggenes, "Yes", "No")
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$Primary.Disease)
    data$G2P_Drug_Gene_Lin <- ifelse(data$Drug_Gene_Lin %in% g2p_dgls, "Yes", "No")
    print("Merged metadata.")
  }
  if(database == "CRISPR") {
    print(paste0("Processing ", database, " data."))
    # Melt CRISPR dataset for merging
    data <- melt(crispr_all, id.vars = "Broad_ID", measure.vars = colnames(crispr_all)[2:ncol(crispr_all)], variable.name = "Gene", value.name = "Score")
    # Merge cell line metadata
    data <- merge(data, select(ccl_meta, Cell_line, Histology, CCLE_Name, Broad_ID, Primary.Disease), by = "Broad_ID", all.x = TRUE)
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$Primary.Disease)
    print("Merged metadata.")
  }

@@

```{r, eval=FALSE}
crispr_all <- read.delim("./data_munging/19Q1/gene_effect_corrected_20190305.csv.gz", sep = ",", header  = TRUE, check.names = FALSE)
# Remove Entrez gene IDs from colnames
colnames(crispr_all) <- gsub(" .*", "", colnames(crispr_all))
colnames(crispr_all)[1] <- "Broad_ID"

@@

# Filter G2P CRISPR for point mutations
crispr_ptmuts <- filter(crispr_data, PointMutation == 1)
crispr_ptmuts <- subset(crispr_ptmuts, select = -c(PointMutation))
# Filter CRISPR for G2P genes in G2P
crispr_g2p <- filter(crispr_ptmuts, Gene %in% g2p_genes)
crispr_g2p <- unique(crispr_g2p)
crispr_g2p$Gene <- factor(crispr_g2p$Gene)
crispr_g2p$Mutation_Status <- factor(crispr_g2p$Mutation_Status)
crispr_g2p$Primary.Disease <- factor(crispr_g2p$Primary.Disease)
```

Test merging and filtering functions on CRISPR data:
```{r, eval=FALSE}
# crispr_all <- read.delim("./data_munging/19Q1/gene_effect_corrected_20190305.csv.gz", sep = ",", header  = TRUE, check.names = FALSE)
# # Remove Entrez gene IDs from colnames
# colnames(crispr_all) <- gsub(" .*", "", colnames(crispr_all))
# colnames(crispr_all)[1] <- "Broad_ID"

```{r}
crispr_fxn <- readRDS("./../htscreens_giant_files/crispr_data_19Q1_g2pgenes_fxn.rds")
crispr_ptmuts_fxn <- readRDS("./../htscreens_giant_files/crispr_data_19Q1_g2pgenes_ptmuts_fxn.rds")
```

# G2P summary figures

@@

```{r fig.height=6, fig.width=12}
g2p_dg_counts <- g2p_all %>% count(Drug, , Gene) %>% filter(!is.na(Drug), !is.na(Gene))
g2p_dg_counts <- dcast(g2p_dg_counts, Gene ~ Drug, value.var = "n", fun.aggregate = mean, fill = 0) %>% `rownames<-`(.[,1]) %>% select(-Gene)
g2p_dg_counts <- g2p_dg_counts[order(rowSums(-g2p_dg_counts)), order(colSums(-g2p_dg_counts))]
g2p_dg_counts[g2p_dg_counts == 0] <- NA

# Define color scale
col_fun <- colorRamp2(c(0, 103), c("mistyrose", "purple4"))
# Legend properties list
ht_legend_list <- list(legend_direction = "horizontal", title_position = "lefttop", legend_width = unit(6, "in"), border = "black")
ht <- Heatmap(g2p_dg_counts, 
              name = "Frequency",
              col = col_fun,
              na_col = "white",
              rect_gp = gpar(col = "gray88"),
              heatmap_legend_param = ht_legend_list,
              cluster_rows = FALSE, 
              cluster_columns = FALSE, 
              row_names_side = "left",
              row_title = "Genes",
              column_title = "Drugs", 
              column_title_side = "bottom")

# png(filename = out_filepath, width = 7.5, height = 8, units = "in", res = 500)
draw(ht, heatmap_legend_side = "top")
# dev.off()
```

@@

@@

Spearman correlations for gene expression and copy number:
```{r}
crispr_spear_ge <- transform(crispr_g2p %>% group_by(Gene) %>% summarize(p = cor.test(y = Score, x = TPM, method = "spearman", use = "complete.obs")$p.value, p_log10 = -log10(p), corr = cor.test(y = Score, x = TPM, method = "spearman", use = "complete.obs")$estimate, Metric = "Gene Expression"), Gene = reorder(Gene, corr))
crispr_spear_cn <- transform(crispr_g2p %>% group_by(Gene) %>% summarize(p = cor.test(y = Score, x = CN, method = "spearman", use = "complete.obs")$p.value, p_log10 = -log10(p), corr = cor.test(y = Score, x = CN, method = "spearman", use = "complete.obs")$estimate, Metric = "Copy Number"), Gene = reorder(Gene, corr))
```

Create CRISPR two-column, KRAS-inset plot for CN and GE:
```{r fig.height=6, fig.width=12}
spearman_KRAS <- rbind(filter(crispr_g2p, Gene == "KRAS") %>% summarize(p = cor.test(y = Score, x = TPM, method = "spearman", use = "complete.obs")$p.value, corr = cor.test(y = Score, x = TPM, method = "spearman", use = "complete.obs")$estimate, Metric = "Gene Expression"), filter(crispr_g2p, Gene == "KRAS") %>% summarize(p = cor.test(y = Score, x = CN, method = "spearman", use = "complete.obs")$p.value, corr = cor.test(y = Score, x = CN, method = "spearman", use = "complete.obs")$estimate, Metric = "Copy Number"))

crispr_ge_spearman_corr_inset <- ggplot(filter(crispr_g2p, Gene == "KRAS"), aes(x = TPM, y = Score)) +
  geom_point(pch = 21, fill = "cadetblue3", alpha = 0.6) +
  labs(x = "KRAS Gene Expression (TPM)", y = "CERES Score", subtitle = paste0("Spearman correlation coefficient = ", round(spearman_KRAS$corr[1], 2), " (p = ", formatC(as.numeric(spearman_KRAS$p[1]), format = "e", digits = 1), ")"))

crispr_cn_spearman_corr_inset <- ggplot(filter(crispr_g2p, Gene == "KRAS"), aes(x = CN, y = Score)) +
  geom_point(pch = 21, fill = "palegreen2", alpha = 0.6) +
  labs(x = expression("KRAS Copy Number"~(log[2]-ratio)), y = "CERES Score", subtitle = paste0("Spearman correlation coefficient = ", round(spearman_KRAS$corr[2], 2), " (p = ", formatC(as.numeric(spearman_KRAS$p[2]), format = "e", digits = 1), ")"))

@@

ggsave("./plots/manuscript/crispr_spearman_corr_plot.png", crispr_spearman_corr_plot, device = "png", dpi = 450, width = 20, height = 10, units = "in")
```

@@

Kolmogorov-Smirnov tests comparing DGLs that are in a G2P DGL and not.

**NEED TO FILTER FOR ONLY G2P DRUG-GENE COMBINATIONS PRIOR TO KS TEST**

```{r, eval=TRUE, fig.height=6, fig.width=12}
# g2p_corr_all <- G2PCorrTable(ccle_ptmuts_df = ccle_ptmuts, ctrp_ptmuts_df = ctrp_ptmuts, gdsc_ptmuts_df = gdsc_ptmuts)
# saveRDS(g2p_corr_all, "./data_munging/rds/g2p_corr_all.rds")
g2p_corr_all <- readRDS("./data_munging/rds/g2p_corr_all.rds")
g2p_mut_stats <- G2PCorrStatsTables(corr_df = g2p_corr_all, table_type = "mutation")

@@

ggsave("./plots/manuscript/g2p_spearman_ks_res_plot.png", g2p_ks_res_plot_spearman, device = "png", dpi = 500, width = 12, height = 6, units = "in")
```

@@

>>>>

    # Get all G2P gene-cell line combinations
    grid <- expand.grid("Cell_line" = unique(filt$Cell_line), "Gene" = g2p_genes)
    data <-  merge(filt, grid, by = "Cell_line", all.y = TRUE)
    print("Merged grid.")
    # Merge metadata
    data <-  merge(data, select(ccl_meta, Cell_line, Histology, CCLE_Name, Broad_ID, Primary.Disease), by = c("Cell_line", "Histology", "CCLE_Name", "Broad_ID"), all.x = TRUE)
    ## Add columns for G2P drug-gene and drug-gene-lineage sorting
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$G2P_Drug_Gene <- ifelse(data$Drug_Gene %in% g2p_druggenes, "Yes", "No")
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$Primary.Disease)
    data$G2P_Drug_Gene_Lin <- ifelse(data$Drug_Gene_Lin %in% g2p_dgls, "Yes", "No")
    print("Merged metadata.")
  }
  if(database == "CRISPR") {
    print(paste0("Processing ", database, " data."))
    # Melt CRISPR dataset for merging
    data <- melt(crispr_all, id.vars = "Broad_ID", measure.vars = colnames(crispr_all)[2:ncol(crispr_all)], variable.name = "Gene", value.name = "Score")
    # Get all G2P gene-cell line combinations
    data <-  filter(data, Gene %in% g2p_genes)
    # Merge cell line metadata
    data <- merge(data, select(ccl_meta, Cell_line, Histology, CCLE_Name, Broad_ID, Primary.Disease), by = "Broad_ID", all.x = TRUE)
    data$Drug_Gene <- paste0(data$Drug, "_", data$Gene)
    data$Drug_Gene_Lin <- paste0(data$Drug, "_", data$Gene, "_", data$Primary.Disease)
    print("Merged metadata.")
  }

@@

Merge all CRISPR genes:
```{r, eval=FALSE}
crispr_all <- read.delim("./data_munging/19Q1/gene_effect_corrected_20190305.csv.gz", sep = ",", header  = TRUE, check.names = FALSE)
# Remove Entrez gene IDs from colnames
colnames(crispr_all) <- gsub(" .*", "", colnames(crispr_all))
colnames(crispr_all)[1] <- "Broad_ID"

@@

# Filter G2P CRISPR for point mutations
crispr_ptmuts <- filter(crispr_data, PointMutation == 1)
crispr_ptmuts <- subset(crispr_ptmuts, select = -c(PointMutation))
crispr_ptmuts <- filter(crispr_ptmuts, Gene %in% g2p_genes)
```

# G2P summary heatmap

@@

```{r fig.height=6, fig.width=8}
g2p_dg_counts <- g2p_all %>% count(Drug, Gene) %>% filter(!is.na(Drug), !is.na(Gene))
g2p_dg_counts <- dcast(g2p_dg_counts, Drug ~ Gene, value.var = "n", fun.aggregate = mean, fill = 0) %>% `rownames<-`(.[,1]) %>% select(-Drug)
g2p_dg_counts <- g2p_dg_counts[order(rowSums(-g2p_dg_counts)), order(colSums(-g2p_dg_counts))]
g2p_dg_counts[g2p_dg_counts == 0] <- NA

# Create annotation layer
hta_data <- data.frame(Drug = rownames(g2p_dg_counts), CCLE = NA, CTRP = NA, GDSC = NA)
hta_data$CCLE <- ifelse(hta_data$Drug %in% unique(ccle_ptmuts$Drug), "Screened", "Not screened")
hta_data$CTRP <- ifelse(hta_data$Drug %in% unique(ctrp_ptmuts$Drug), "Screened", "Not screened")
hta_data$GDSC <- ifelse(hta_data$Drug %in% unique(gdsc_ptmuts$Drug), "Screened", "Not screened")
rownames(hta_data) <- hta_data$Drug
hta_data$Drug <- NULL

hta_annot <- HeatmapAnnotation(text = anno_text(colnames(hta_data), offset = unit(0.1, "in"), gp = gpar(fontsize = 10)))
hta_legend_list <- list(legend_direction = "horizontal", border = "gray70", nrow = 2)
hta <- Heatmap(hta_data,
               name = "Functional screen status           ",
               col = c("Screened" = "purple4", "Not screened" = "gray97"), 
               rect_gp = gpar(col = "gray70"),
               top_annotation = hta_annot,
               heatmap_legend_param = hta_legend_list,
               row_names_side = "left",
               row_names_gp = gpar(fontsize = 10),
               column_names_side = "top",
               column_names_gp = gpar(fontsize = 10),
               show_column_names = FALSE,
               cluster_rows = FALSE, 
               cluster_columns = FALSE,
               row_title = "Drugs",
               row_title_gp = gpar(fontface = "bold", fontsize = 14),
               column_title = "Dataset",
               column_title_gp = gpar(fontface = "bold", fontsize = 14),
               width = unit(1.5, "in"))

# Define color scale
col_fun <- colorRamp2(c(0, 100), c("mistyrose", "purple4"))
# Legend properties list
ht_legend_list <- list(legend_direction = "horizontal", legend_width = unit(2.5, "in"), border = "gray70")
ht <- Heatmap(g2p_dg_counts, 
              name = "Number of G2P interpretations",
              col = col_fun,
              na_col = "gray97",
              rect_gp = gpar(col = "gray70"),
              heatmap_legend_param = ht_legend_list,
              cluster_rows = FALSE, 
              cluster_columns = FALSE, 
              show_row_names = FALSE,
              row_names_side = "left",
              row_names_gp = gpar(fontsize = 10),
              column_names_side = "bottom",
              column_names_gp = gpar(fontsize = 10),
              row_title = "Drugs",
              row_title_gp = gpar(fontface = "bold", fontsize = 14),
              column_title = "Genes", 
              column_title_gp = gpar(fontface = "bold", fontsize = 14),
              column_title_side = "bottom")

# png(filename = "./plots/manuscript/g2p_summary_heatmap.png", width = 12, height = 10, units = "in", res = 500)
draw(hta + ht, gap = unit(5, "mm"), heatmap_legend_side = "top")
# dev.off()
```

@@

## Plots

@@

Spearman correlations for gene expression and copy number:
```{r, warning=FALSE}
crispr_spear_ge <- transform(crispr_ptmuts %>% group_by(Gene) %>% summarize(p = cor.test(y = Score, x = TPM, method = "spearman", use = "complete.obs")$p.value, p_log10 = -log10(p), corr = cor.test(y = Score, x = TPM, method = "spearman", use = "complete.obs")$estimate, Metric = "Gene Expression"), Gene = reorder(Gene, corr))
crispr_spear_cn <- transform(crispr_ptmuts %>% group_by(Gene) %>% summarize(p = cor.test(y = Score, x = CN, method = "spearman", use = "complete.obs")$p.value, p_log10 = -log10(p), corr = cor.test(y = Score, x = CN, method = "spearman", use = "complete.obs")$estimate, Metric = "Copy Number"), Gene = reorder(Gene, corr))
```

## Plots

Create CRISPR two-column, KRAS-inset plot for CN and GE:
```{r fig.height=6, fig.width=12}
spearman_KRAS <- rbind(filter(crispr_ptmuts, Gene == "KRAS") %>% summarize(p = cor.test(y = Score, x = TPM, method = "spearman", use = "complete.obs")$p.value, corr = cor.test(y = Score, x = TPM, method = "spearman", use = "complete.obs")$estimate, Metric = "Gene Expression"), filter(crispr_ptmuts, Gene == "KRAS") %>% summarize(p = cor.test(y = Score, x = CN, method = "spearman", use = "complete.obs")$p.value, corr = cor.test(y = Score, x = CN, method = "spearman", use = "complete.obs")$estimate, Metric = "Copy Number"))

crispr_ge_spearman_corr_inset <- ggplot(filter(crispr_ptmuts, Gene == "KRAS"), aes(x = TPM, y = Score)) +
  geom_point(pch = 21, fill = "cadetblue3", alpha = 0.6) +
  labs(x = "KRAS Gene Expression (TPM)", y = "CERES Score", subtitle = paste0("Spearman correlation coefficient = ", round(spearman_KRAS$corr[1], 2), " (p = ", formatC(as.numeric(spearman_KRAS$p[1]), format = "e", digits = 1), ")"))

crispr_cn_spearman_corr_inset <- ggplot(filter(crispr_ptmuts, Gene == "KRAS"), aes(x = CN, y = Score)) +
  geom_point(pch = 21, fill = "palegreen2", alpha = 0.6) +
  labs(x = expression("KRAS Copy Number"~(log[2]-ratio)), y = "CERES Score", subtitle = paste0("Spearman correlation coefficient = ", round(spearman_KRAS$corr[2], 2), " (p = ", formatC(as.numeric(spearman_KRAS$p[2]), format = "e", digits = 1), ")"))

@@

# ggsave("./plots/manuscript/crispr_spearman_corr_plot.png", crispr_spearman_corr_plot, device = "png", dpi = 450, width = 20, height = 10, units = "in")
```

@@

Kolmogorov-Smirnov tests comparing DGLs that are in a G2P DGL and not.
```{r, eval=TRUE, fig.height=6, fig.width=12}
# g2p_corr_all <- G2PCorrTable(ccle_ptmuts_df = ccle_ptmuts, ctrp_ptmuts_df = ctrp_ptmuts, gdsc_ptmuts_df = gdsc_ptmuts)
# saveRDS(g2p_corr_all, "./data_munging/rds/g2p_corr_all.rds")
g2p_corr_all <- readRDS("./data_munging/rds/g2p_corr_all.rds")
g2p_mut_stats <- G2PCorrStatsTables(corr_df = g2p_corr_all, table_type = "mutation")

@@

# ggsave("./plots/manuscript/g2p_spearman_ks_res_plot.png", g2p_ks_res_plot_spearman, device = "png", dpi = 500, width = 12, height = 6, units = "in")
```

@@

Test filtering CRISPR data for G2P genes using function:
```{r, eval=FALSE}
crispr_all <- read.delim("./data_munging/19Q1/gene_effect_corrected_20190305.csv.gz", sep = ",", header  = TRUE, check.names = FALSE)
# Remove Entrez gene IDs from colnames
colnames(crispr_all) <- gsub(" .*", "", colnames(crispr_all))
colnames(crispr_all)[1] <- "Broad_ID"

```{r, eval=FALSE}
crispr_fxn <- readRDS("./../htscreens_giant_files/crispr_data_19Q1_g2pgenes_fxn.rds")
crispr_ptmuts_fxn <- readRDS("./../htscreens_giant_files/crispr_data_19Q1_g2pgenes_ptmuts_fxn.rds")
```
